Inequality in non-cognitive skills: a cross-cohort comparison

### 2018-07-31 codes/R/notes/paper_jpube_partinv.Rnw
with partial invariance

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_selectvars.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/9_gelbfig.R"))
read_chunk(here("codes/R/getpars.R"))

@@

@@

@@

@@

\documentclass[pdftex,10pt,twoside,a4paper]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amstext}% for resizing the symbol in math
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{booktabs}
\usepackage{tocloft}
\usepackage{dsfont}
\usepackage{xcolor}
\usepackage{accents}
\usepackage{rotating}
\usepackage{pdflscape}  % for landscape page
\usepackage{lmodern}  %font
\usepackage{natbib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{tikz}
\usepackage{bm}
\usepackage{array}
\usepackage{multirow}
\usepackage[multiple]{footmisc}   % for multiple footnotes
\usepackage{textgreek}            % for greek letters in text
\usepackage{ragged2e}             % for table notes justification
\usepackage{setspace} % for spacing
\onehalfspacing

@@

% page setup (from Racine)
\setlength{\topmargin}{0cm}
\setlength{\textheight}{24cm}
\setlength{\oddsidemargin}{.0cm}
\setlength{\evensidemargin}{.0cm}
\setlength{\textwidth}{16.5cm}
\setlength{\footskip}{1.5cm}
\setlength{\footnotesep}{0.4cm}
\linespread{1.3}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

table <- rbind(afitab[[1]],afitab[[2]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
                    digits = 4
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3)),
                              command = c(
                              # header
            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
            Model & Num. params & $\\chi^2$ & RMSEA & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSEA & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
                       \\textbf{A: All ages} \\\\",
                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSEA = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat. Panel A shows results for the whole sample of children in the BCS and MCS cohorts. Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts -- 59-61 months.
\par}
\end{table}
\end{landscape}

>>>>

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_part.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_selectvars.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/9_gelbfig.R"))
read_chunk(here("codes/R/getpars.R"))

@@

<<FUN_GETPARS>>=
# load function that gets parameters
@

@@

<<FA_INV_PART>>=
# partial invariance (many models)
@

@@

<<FA_SCORES_MAIN>>=
# produce scores for main model
@

<<FA_SCORES_BIND>>=
# bind scores with data
@

<<FA_SCORES_ALT>>=
# produce scores for alternative anchoring model
@

@@

<<FACINEQ_MAKE>>=
# make factor inequality graphs
@

\documentclass[pdftex,10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amstext}% for resizing the symbol in math
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{booktabs}
\usepackage{tocloft}
\usepackage{dsfont}
\usepackage{xcolor}
\usepackage{accents}
\usepackage{rotating}
\usepackage{pdflscape}  % for landscape page
\usepackage{lmodern}  %font
\usepackage{natbib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{tikz}
\usepackage{bm}
\usepackage{array}
\usepackage{multirow}
\usepackage[multiple]{footmisc}   % for multiple footnotes
\usepackage{textgreek}            % for greek letters in text
\usepackage{ragged2e}             % for table notes justification
\usepackage{setspace} % for spacing
\onehalfspacing

@@

% page setup (from Racine)
\setlength{\topmargin}{0cm}
\setlength{\textheight}{23cm}
\setlength{\oddsidemargin}{.0cm}
\setlength{\evensidemargin}{.0cm}
\setlength{\textwidth}{16.5cm}
\setlength{\footskip}{1.5cm}
\setlength{\footnotesep}{0.4cm}

@@

% ADD PARTIAL INVARIANCE

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

table <- rbind(fit_main,fit_agesub)

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
                    digits = 4
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,4)),
                              command = c(
                              # header
            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
            Model & Num. params & $\\chi^2$ & RMSEA & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSEA & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
                       \\textbf{A: All ages} \\\\",
                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSEA = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat. Panel A shows results for the whole sample of children in the BCS and MCS cohorts. Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
\par}
\end{table}
\end{landscape}

\clearpage
\begin{table}
\caption{Estimated parameters -- Threshold and loading invariance model (anchored)}\label{tab:measpars}
\centering\setstretch{1.2}\scriptsize
<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=

table <- getmeaspars(fa.tlmt[[1]], groups = 4, mode = "tl")

# sequences to add rows
rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
rwsl <- as.list(rws)

print.xtable(xtable(table,
              align="C{0cm}C{.06\\textwidth}C{.04\\textwidth}C{.07\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}",
              digits = 3
                    ),
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                          " &
                                            & \\textbf{Loadings}
                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
                                            \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}

                                            &
                                            & $\\lambda$
                                            & $\\tau_1$ & $\\tau_2$
                                            & \\multicolumn{3}{c}{$\\nu$}
                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\

                                            Measure
                                            & Factor
                                            & All
                                            & All & All
                                            & BCS F & MCS M & MCS F
                                            & BCS F & MCS M & MCS F
                                            \\\\ "
                      )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)
@
\end{table}

\begin{table}
\caption{Estimated latent variable parameters -- Threshold and loading invariance model (anchored)}\label{tab:lvpars}
\centering\setstretch{1.2}\scriptsize
<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
require(xtable)

lvtab <- getlvpars(fa.tlmt[[1]], groups = 4)
table <- rbind(lvtab$Males,lvtab$Females)

print.xtable(xtable(table,
                     align="C{0cm}L{.13\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 3
                    ),
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,2,4)),
                              command = c(
                              # header
                      "\\toprule
                      & \\multicolumn{4}{c}{\\textbf{BCS}}
                      & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}

                      & Mean
                      & \\multicolumn{2}{c}{Covariance}
                      & \\multicolumn{1}{c}{Correlation}
                      & Mean
                      & \\multicolumn{2}{c}{Covariance}
                      & \\multicolumn{1}{c}{Correlation} \\\\

                      \\textbf{Males} \\\\",
                      "\\textbf{Females} \\\\",
                      "\\bottomrule"
                                    )),
             hline.after = NULL,  # remove default booktabs lines
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage

<<FA_FIT_PART, dependson="FA_INV_PART", results='markup'>>=
head(partialfit[order(partialfit$rmsea),], 15)
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
@
\caption{Scores}
\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 1, 6, 9, 10.
\par}
\end{figure}

\begin{figure}
<<FACDENS_ALT, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
# graph x axis boundaries
maxx <- max(scores2plot_alt[names(scores2plot_alt) %in% c("EXT","INT")]) +.05
minx <- min(scores2plot_alt[names(scores2plot_alt) %in% c("EXT","INT")]) -.05

# common options
addopts.dens <- function(x) {
  x <- x +
    theme(
      axis.title.y=element_blank(),
      legend.position="none"
    ) +
    scale_x_continuous(name = "Score", breaks = seq(-3.5,1.5,.5)) +
    coord_cartesian(xlim = c(minx, maxx), ylim = c(0,.62)) +
    scale_fill_discrete("") + # remove fill guide title
    scale_colour_discrete(guide=FALSE) +  # remove colour legend
    geom_density(alpha = 0.1, bw = "nrd", adjust=1.5, kernel = "epanechnikov")
  return(x)
}

# densities of factor scores
pdext.ebm.m <- ggplot(subset(scores2plot_alt, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Males)")
pdint.ebm.m <- ggplot(subset(scores2plot_alt, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Males)")
pdext.ebm.f <- ggplot(subset(scores2plot_alt, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Females)")
pdint.ebm.f <- ggplot(subset(scores2plot_alt, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Females)")
denslist <- list(pdext.ebm.m, pdint.ebm.m, pdext.ebm.f, pdint.ebm.f) 
denslist <- lapply(denslist, addopts.dens) # apply options to all graphs
pcol <- plot_grid( denslist[[1]], denslist[[3]], denslist[[2]], denslist[[4]],
                   align = 'vh',
                   hjust = -1,
                   nrow = 2
)
# add legend
legend_b <- get_legend(denslist[[1]] + theme(legend.position="bottom"))
p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
p

@
\caption{Scores -- alternative restrictions}
\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 3, 6, 8, 11.
\par}
\end{figure}

\begin{figure}
<<FACDENS_ALT2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
# graph x axis boundaries
maxx <- max(scores2plot_alt2[names(scores2plot_alt2) %in% c("EXT","INT")]) +.05
minx <- min(scores2plot_alt2[names(scores2plot_alt2) %in% c("EXT","INT")]) -.05

# common options
addopts.dens <- function(x) {
  x <- x +
    theme(
      axis.title.y=element_blank(),
      legend.position="none"
    ) +
    scale_x_continuous(name = "Score", breaks = seq(-3.5,1.5,.5)) +
    coord_cartesian(xlim = c(minx, maxx), ylim = c(0,.62)) +
    scale_fill_discrete("") + # remove fill guide title
    scale_colour_discrete(guide=FALSE) +  # remove colour legend
    geom_density(alpha = 0.1, bw = "nrd", adjust=1.5, kernel = "epanechnikov")
  return(x)
}

# densities of factor scores
pdext.ebm.m <- ggplot(subset(scores2plot_alt2, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Males)")
pdint.ebm.m <- ggplot(subset(scores2plot_alt2, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Males)")
pdext.ebm.f <- ggplot(subset(scores2plot_alt2, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Females)")
pdint.ebm.f <- ggplot(subset(scores2plot_alt2, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Females)")
denslist <- list(pdext.ebm.m, pdint.ebm.m, pdext.ebm.f, pdint.ebm.f) 
denslist <- lapply(denslist, addopts.dens) # apply options to all graphs
pcol <- plot_grid( denslist[[1]], denslist[[3]], denslist[[2]], denslist[[4]],
                   align = 'vh',
                   hjust = -1,
                   nrow = 2
)
# add legend
legend_b <- get_legend(denslist[[1]] + theme(legend.position="bottom"))
p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
p

@
\caption{Scores -- alternative restrictions}
\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 3, 6, 9, 10.
\par}
\end{figure}

\begin{figure}
<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
# plot
fi <- fi_fsc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Inequality by father occupation -- normalised}
\end{figure}

\begin{figure}
<<FACINEQ_FSC2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
# plot
fi <- fi_fsc2 # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Inequality by father occupation -- not normalised}
\end{figure}

\begin{figure}
<<FACINEQ_FSC3, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
fi_fsc3 <- plotineq(scores2plot_alt, "fscl5wb", ylab = "Score ", xlab = "Father Occupation (5)",
                   ylim = c(-.8,1.0))
fi <- fi_fsc3 # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Inequality by father occupation -- not normalised, alternative restrictions}
\end{figure}

\begin{figure}
<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
# plot
fi <- fi_ps # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Inequality by mother education -- normalised}
\end{figure}

\begin{figure}
<<FACINEQ_PS2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
# plot
fi <- fi_ps2 # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Inequality by mother education -- not normalised}
\end{figure}
